Exploring Compiler Optimization Techniques

### 2014-11-13 doc/paper.tex
Added diagrams and expanded on appendix

\documentclass[12pt, a4paper]{article}
\usepackage[T1]{fontenc}
\usepackage{amsmath, listingsutf8, mathtools}
\usepackage{hyperref}
\usepackage{breakurl}

%\hypersetup{}
%\lstset{numbers=left}
\title{Exploring Compiler Optimization Techniques}
\author{A. D. Blom}
\date{}

@@

Code optimization is most often done by first translating the source code
(as read by a \textit{lexer} and interpreted by a \textit{parser}) into an
\textit{intermediate form}, which is subsequently converted into assembly.\cite{dragon}
Such an intermediate form must be designed in such a way that it can be analyzed and
optimized easily. Such a intermediate form commonly consists of a
list of instructions (using so-called \textit{three-way address code}) grouped
into \textit{basic blocks}. Each basic block has one starting point and one
end-point. The endpoint is always a \textit{terminal instruction}
which causes control flow to leave the block (\verb+jmp+, \verb+split+, \verb+ret+
and \verb+leave+ in the intermediate form used here).

@@

  \section{Register allocation}
  \subsection{As an optimization}
It's hard to categorize register allocation as either an optimization or a
general method used in assembly generation. For the purposes of this article
it'll be classified as the former, because it is a way to speed up generated
assembly.

\subsection{Problem}
The original i386 processors had only very few general purpose registers
(\verb+eax+, \verb+ebx+, \verb+ecx+, \verb+edx+, \verb+esi+ and
\verb+edi+).\cite[section~3.4.1]{intel}

@@

  \bibitem{intel}
  \emph{Intel\textregistered{} 64 and IA-32 Architectures Software Developer's
Manual Volume~1}.
  (2014)

@@

  \subsection{Object oriented programming}
Although the C language doesn't natively feature object oriented syntax, it doesn't
exclude the possibility of writing clean object oriented code. For instance, the
following hierarchy:

INSERT DIAGRAM

Could well be implemented in C as follows:

@@

The following intermediate code is internally represented through a few different
data structures:

@@

@@

\begin{lstlisting}
struct itm_instr {
	struct itm_expr base;

	itm_instr_id_t id;
	struct itm_block *block;
	struct list *operands;
	struct itm_instr *prev, *next;
};
\end{lstlisting}

@@

	/* value, sharing memory */
	union {
		long long i;
		double f;
	} value;
};

>>>>

\documentclass[12pt, a4paper]{article}
\usepackage[T1]{fontenc}
\usepackage{amsmath, listingsutf8, mathtools}
\usepackage{hyperref}
\usepackage{breakurl}
\usepackage{graphicx}
%\hypersetup{}
%\lstset{numbers=left}
\title{Exploring Compiler Optimization Techniques}
\author{A. D. Blom}
\date{}

@@

Code optimization is most often done by first translating the source code
(as read by a \textit{lexer} and interpreted by a \textit{parser}) into an
\textit{intermediate form}, which is subsequently converted into
assembly.\cite[chapter~2]{dragon}
Such an intermediate form must be designed in such a way that it can be analyzed and
optimized easily. Such a intermediate form commonly consists of a
list of instructions (using so-called \textit{three-way address code}) grouped
into \textit{basic blocks}. Each basic block has one starting point and one
end-point. The endpoint is always a \textit{terminal instruction}
which causes control flow to leave the block (\verb+jmp+, \verb+split+, \verb+ret+
and \verb+leave+ in the intermediate form used here).

@@

  \section{Register allocation}
  \label{sec:regalloc}
  \subsection{As an optimization}
It's hard to categorize register allocation as either an optimization or a
general method used in assembly generation. For the purposes of this article
it'll be classified as the former, because it is a way to speed up generated
assembly.

\subsection{Problem}
CPUs have only a limited amount of registers. The original i386 processors had
only six general purpose registers for example (\verb+eax+, \verb+ebx+, \verb+ecx+,
\verb+edx+, \verb+esi+ and \verb+edi+).\cite[section~3.4.1]{intel}

@@

  \bibitem{intel}
  Intel Corporation.
  \emph{Intel\textregistered{} 64 and IA-32 Architectures Software Developer's
Manual Volume~1}.
  (2014)

@@

  \subsection{Object oriented programming}
Although the C language doesn't natively feature object oriented syntax, it doesn't
exclude the possibility of writing clean object oriented code.

\begin{figure}[h]
\begin{center}
\includegraphics[resolution=256]{ABC.png}
\caption{Class diagram example}
\label{fig:abc}
\end{center}
\end{figure}

The class diagram described in figure~\ref{fig:abc} could be implemented in
C as follows:

@@

A variety of structures is needed to store the following snippet internally:

@@

And AST capable of storing such an IR, must be able to store functions, global
variables (unimplemented as of yet), basic blocks, instructions, parameters
(unimplemented as of yet), literals and \verb+undef+ constants. These elements
are implemented through a structure system as described in figure~\ref{fig:ast}.

\begin{figure}[h]
\begin{center}
\includegraphics[resolution=150]{ast.png}
\caption{The AST class diagram}
\label{fig:ast}
\end{center}
\end{figure}

A common base type for most AST elements is \verb+struct itm_expr+ (the
expression base type). It contains a destructor, a (C) typename, and a list of
\textit{tags}. Tags are used to store expression attributes (\textit{"tag"}).
Tags are used in section~\ref{sec:regalloc} for example.

\begin{lstlisting}
struct itm_expr {
	/* derived type identifier
	 *
	 * The enumeration contains for
	 * example:
	 *  ITME_INSTRUCTION,
	 *  ITME_LITERAL
	 *  ITME_BLOCK
	 *  ...
	 */
	enum itm_expr_type etype;

	/* expression's C typename */
	struct ctype *type;
	/* tag list */
	struct list *tags;

	/* destructor
	 * (implemented by derived class) */
	void (*free)(struct itm_expr *e);
	/* to_string for file dumps
	 * (implemented by derived class) */
	void (*to_string)(FILE *f, struct itm_expr *e);
};
\end{lstlisting}

@@

\begin{lstlisting}
struct itm_instr {
	/* expression base type */
	struct itm_expr base;

	/* instruction identifier (add, sub,
	 * leave, etc.) */
	itm_instr_id_t id;
	/* parent block */
	struct itm_block *block;
	/* instruction operand list */
	struct list *operands;
	/* previous and next instructions */
	struct itm_instr *prev, *next;
};
\end{lstlisting}

@@

	/* value, sharing memory */
	union {
		long long i;
		double d;
		float f;
	} value;
};
